---
title: "Welcome and Review"
subtitle: "Stat 230: Applied Regression Analysis"
format: 
  revealjs:
    theme: [serif, styles.scss]
editor: source
---

```{r}
#| include: false
data("bikes", package = "bayesrules")
library(ggformula)

# Define ggplot2 theme with background color
theme_classic <- function() {
  ggplot2::theme_classic() +
    ggplot2::theme(
      plot.background = ggplot2::element_rect(fill = "#fffef5", color = NA),
      panel.background = ggplot2::element_rect(fill = "#fffef5", color = NA),
      axis.line = ggplot2::element_line(color = "#003069"),
      axis.text = ggplot2::element_text(color = "#003069"),
      axis.ticks = ggplot2::element_line(color = "#003069"),
      axis.title = ggplot2::element_text(color = "#003069")
    )
}

# set ggplot2 theme
ggplot2::theme_set(theme_classic())
```

# [PDF version of slides](pdf_slides/01-slides-slr-review.pdf)

## Welcome 👋

-   I'm Adam (he/him)
-   I teach statistics & data science
-   I'm interested in statistics education, data visualization, and R programming

```{r}
#| include: false
#| cache: false
#| eval: false

library(tidyverse)     # ggplot, dplyr, and friends
library(sf)            # Handle spatial data in R in a tidy way
library(tigris)        # Access geographic data from the US Census
library(tidygeocoder)  # Automated geocoding
library(osrm)          # Access OSRM through R
library(ggrepel)       # Nicer non-overlapping labels
library(glue)          # Easier string interpolation
library(scales)        # Nicer labeling functions
library(patchwork)     # Combine plots nicely
library(ggspatial)     # Nicer map features like scale bars

# Custom ggplot theme to make pretty plots
# Get the font at https://fonts.google.com/specimen/Overpass
theme_roadtrip <- function() {
  theme_void(base_family = "Overpass Light") +
    theme(
      plot.title = element_text(family = "Overpass", face = "bold", hjust = 0.5),
      strip.text = element_text(
        family = "Overpass ExtraBold", face = "plain",
        size = rel(1.1), hjust = 0.5),
      plot.background = ggplot2::element_rect(fill = "#fffef5", color = NA),
      panel.background = ggplot2::element_rect(fill = "#fffef5", color = NA)
    )
}


us_states <- states(resolution = "20m", year = 2022, cb = TRUE)

lower_48 <- us_states %>% 
  filter(!(NAME %in% c("Alaska", "Hawaii", "Puerto Rico")))

summer_trips <- tribble(
  ~city, ~state,
  "Northfield", "MN",
  "Fish Creek", "WI",
  "Ames", "IA",
  "Nashville", "TN", 
  "Ocean City", "MD"
)

trips_geocoded <- summer_trips %>% 
  geocode(city = city, state = state, method = "osm") %>%
  st_as_sf(coords = c("long", "lat"), crs = st_crs("EPSG:4326"))
```

```{r}
#| echo: false
#| message: false
#| fig.height: 3
#| fig.width: 5
#| eval: false
ggplot() +
  geom_sf(data = lower_48) +
  geom_sf(data = trips_geocoded) +
  geom_label_repel(
    data = trips_geocoded,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates", seed = 1234,
    size = 3, min.segment.length = 0
  ) +
  coord_sf(crs = st_crs("ESRI:102003")) +  # Albers
  theme_roadtrip()

```


## Your turn

-   Form groups based on value (2, 3, 4...) of the card dealt

-   Introduce yourself to your group

-   Share at least one thing other than just your name and major


# Hi, I'm Emerson 👋





# Simple Linear Regression Review

## Regression in intro {.nostretch}

::::: columns
::: {.column width="60%"}
<br>

```{r}
#| fig.height: 3.5
#| fig.width: 6
#| fig.align: center
#| out.width: 100%
gf_point(rides ~ temp_actual, data = bikes, xlab = "Temperature (F)", ylab = "No. rides", color = "#003069") |>
  gf_lm(rides ~ temp_actual, linewidth = 3, color = "#f15a31") 
```
:::

::: {.column width="40%"}
-   Describe the scatterplot
-   Write the equation of the regression line
-   Interpret the slope and intercept
-   Make a prediction
:::
:::::

## Your turn

-   Work with your group

-   If your card's suit is clubs, you're the designated speaker for your group (be ready!)

-   Work through the review questions

-   Write your group's answer to each question on the whiteboard

## Regression in Stat 230

$y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$ where $\varepsilon_i \overset{\rm iid}{\sim} N (0, \sigma^2)$

```{r}
#| fig.height: 3.5
#| fig.width: 6
#| fig.align: center
library(ggdist)

bike_lm <- lm(rides ~ temp_actual, data = bikes)
bike_dsn <- data.frame(
  temp_actual = c(50, 60, 70, 80),
  yhat = predict(bike_lm, data.frame(temp_actual = c(50, 60, 70, 80))),
  resid_sd = rep(sigma(bike_lm), 4)
)


ggplot(data = bikes, aes(x = temp_actual, y = rides)) +
  geom_point(alpha = 0.4, color = "#003069") +
  geom_lm(color = "#f15a31") +
  stat_slab(aes(x = temp_actual, ydist = "norm", arg1 = yhat, arg2 = resid_sd), data = bike_dsn, alpha = 0.4, inherit.aes = FALSE, scale = 0.6, fill = "#f15a31") +
  theme_classic() +
  xlab("Temperature") +
  ylab("No. rides")
```

## Simple Linear Regression Model

$y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$ where $\varepsilon_i \overset{\rm iid}{\sim} N (0, \sigma^2)$

<br>

-   Linear relationship between $x$ and $y$
-   Errors are independent and identically distributed (iid)
-   Errors are normally distributed
-   Errors have mean 0
-   Variance of the errors doesn't depend on $x$

## Notation

Mean function:

$\quad E(Y|X)=\mu\{Y|X\} = \beta_0 + \beta_1 x_i$

<br>

Fitted model equation:

$\quad \widehat{y}_i = \widehat{\mu}\{Y|X\} = \widehat{\beta}_0 + \widehat{\beta}_1 x_i$

# Overview of Stat 230

## Overview of the term

Goal: Develop models to answer research questions

```{mermaid}
timeline
%%{init: {'theme':'neutral'}}%%
    
    Weeks 1-2 : Simple linear regression
           : Review
           : Inference
           : Model diagnostics
           
    Weeks 3-5 : Multiple linear regression
           : Interpretation
           : Inference
           : Model diagnostics
           : Model selection
           
    Weeks 6-8 : Logistic regression
           : Interpretation
           : Inference
           : Model diagnostics
           : Model selection
           
    Weeks 9-10 : Poisson regression
           : Interpretation
           : Inference
```

## Typical week

:::::{.r-fit-text}
:::: {.r-hstack}
::: { style="background: #ffd24f; width: 33%; height: 90%; margin: 20px; padding: 0.5em; border-radius: 5px;"}

**Monday**

- Pre-class reading/video
- Reflection questions

<br>

- Class meeting
- Work on problems
:::

::: { style="background: #ffd24f; width: 33%; height: 90%; margin: 20px; padding: 0.5em; border-radius: 5px;"}
**Wednesday**

- Pre-class reading/video
- Reflection questions

<br>

- Class meeting
- Work on problems
:::

::: { style="background: #ffd24f; width: 33%; height: 90%; margin: 20px; padding: 0.5em; border-radius: 5px;"}

**Friday**

- Pre-class reading/video
- Reflection questions
- Homework due by the start of class
- Class meeting
- Work on problems
:::
::::
:::::

## Tools

::::{.columns}
::: {.column width="30%"}

**Moodle**

<br>
<br>

**Gradescope**

<br>


**R and RStudio**

<br>

**R Markdown**

:::
::: {.column width="70%"}
Class website; look here for all materials, links, etc.

<br>

Submit assignments, get feedback

<br>


Our computational engine

<br>

Dynamic documents for assignments, case studies, projects
:::
::::

# Before next class...

1.  Explore the Moodle page
2.  Read the syllabus -- if you have questions, ask them on the syllabus!
3.  Review hypothesis tests and confidence intervals as necessary (the basic ideas)
3.  Read section 7.4 of Sleuth
4.  Complete the pre-class questions (bring your answers to class)
