---
title: "Model Validation"
subtitle: "Stat 230: Applied Regression Analysis"
format: 
  revealjs:
    theme: [serif, styles.scss]
    scrollable: true
editor: source
editor_options: 
  chunk_output_type: console
code-annotations: hover
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = TRUE, dev = 'svg', comment = NULL)

library(ggformula)
library(dplyr)
library(broom)
library(gridExtra)
library(effects)
library(ggthemes)
library(car)
library(here)
library(gt)
library(gtsummary)
library(gtsummary)
library(DAAG)
library(GGally)
library(ggeffects)
library(Sleuth3)

options(digits=4) # display four significant digits by default

theme_classic <- function() {
  ggplot2::theme_classic() +
    ggplot2::theme(
      plot.background = ggplot2::element_rect(fill = "#fffef5", color = NA),
      panel.background = ggplot2::element_rect(fill = "#fffef5", color = NA),
      axis.line = ggplot2::element_line(color = "#003069"),
      axis.text = ggplot2::element_text(color = "#003069"),
      axis.ticks = ggplot2::element_line(color = "#003069"),
      axis.title = ggplot2::element_text(color = "#003069")
    )
}

graph_paper <- ggplot(data = data.frame(x = 0:20, y = 0:20), aes(x, y)) +
  theme_classic() +
  scale_x_continuous(name = "Age", breaks = seq(0, 20, by = 2), expand = c(0, 0)) +
  scale_y_continuous(name = "FEV", breaks = seq(0, 20, by = 2), expand = c(0, 0)) +
  theme(  
    panel.grid.major = element_line(color = "gray60", size = 0.5), # Major grid lines
    panel.grid.minor = element_line(color = "gray60", size = 0.25), # Minor grid lines
  )

# set ggplot2 theme
ggplot2::theme_set(theme_classic())

library(fivethirtyeight)
```


# First, other model comparison metrics

## Metrics that avoid testing

- Adjusted $R^2$

- Akaike’s Information Criteria (AIC)

- Bayesian Information Criteria (BIC)

- Mallow's $C_p$

These do not rely on p-values, and they balance both complexity and fit

## Mallow's $C_p$

Let $p$ be # of coefficients for the model in question

$$C_p = \underbrace{\frac{SSE_{p}}{MSE_{full}}}_{\text{error part}} - \underbrace{(n - 2p)}_{\text{penalty for complexity}}$$

- Want small $C_p$, but also want $C_p \approx p$

- $C_m = m$ for the biggest model ( $m$ = # of coefficients)

- $C_p$ focuses on prediction

- Or, $C_p$ minimizes $SSE + \text{penalty}$



## AIC

$$AIC = \underbrace{n \log(SSE/n)}_{\text{error part}} + \underbrace{2p}_{\text{penalty for complexity}}$$

- choose model with smaller AIC

- models do not need to be nested to be compared

- based on asymptotic theory

- generally favors models that are bigger than the "true" model



## BIC

$$BIC = \underbrace{n \log(SSE/n)}_{\text{error part}} + \underbrace{p \log(n)}_{\text{penalty for complexity}}$$

- choose model with smaller BIC

- models do not need to be nested to be compared

- based on asymptotic theory

- usually larger penalty than AIC $\Longrightarrow$ leads to smaller models


# Model validation

## Example: Candy rankings

- "we devised an experiment: Pit dozens of fun-sized candy varietals against one another"
- Crowd-sourced evaluations
- response = win percentage
- possible predictor variables:

:::columns
:::{.column width=50%}
Indicator variables:
chocolate, fruity, caramel,
peanutyalmondy, 
nougat,
crispedricewafer, 
hard, 
bar,
pluribus (one of many candies)
:::
:::{.column width=50%}
Quantitative variables:
sugarpercent, <br>
pricepercent
:::
:::

:::{.footer}
https://fivethirtyeight.com/videos/the-ultimate-halloween-candy-power-ranking/
:::

## Selected model

Starting with a model that contained all predictor variables, dropped variables that were unimportant:

```{r}
#| cache: true
#| include: false
set.seed(1234)
nsim <- 100
coef_list <- vector("list", nsim)
y_list <- vector("list", nsim)
for(i in seq_len(nsim)) {
  candy <- candy_rankings
  candy$winpercent <- 100 * runif(nrow(candy))
  model <- step(lm(winpercent ~ . - competitorname, data = candy), k = 2, trace = FALSE)
  coef_list[[i]] <- broom::tidy(model)
  y_list[[i]] <- candy$winpercent
}
```

<br>

```{r}
#| echo: false
candy_fake <- candy
candy_fake$winpercent <- y_list[[3]]
selected_model <- lm(winpercent ~ caramel + peanutyalmondy + hard + bar, data = candy_fake)

tidy(selected_model) |>
mutate(p.value = style_pvalue(p.value)) |>
  gt() |> 
  fmt_number(columns = c(estimate, std.error, statistic), decimals = 3) |>
  fmt_number(columns = p.value, decimals = 4) |>
  tab_options(
    table.font.size = px(36),
    table.background.color = "#fffef5",
    table.font.color = "#003069"
  )
```


## But wait!

- I replaced the real win percentage with randomly generated values between 0 and 100!

- We still got "significant" results...


## Beware of inference after model selection!

If we use the same data set to conduct inference as we used to select our model we run into (related) issues 

- overfitting

- selection bias

- under-estimation of MSE


## Validation

If you want to run inference on variables used in model selection...

1. Collect new data for inference -> might be possible in controlled experiments

2. Compare results with past results

3. Use a holdout sample -> most practical


## Data splitting

- Split your original data set into two pieces: a **training set** and a **validation set**

- Training set is only used for model selection, should have at least 6 to 10 times as many rows as there are slope coefficients in the biggest model considered

- What split?

    + Random 50/50 split is a starting point
    + Increase size of training set until you get 6-10 x rows as slopes
    
- Sometimes you don't have enough data to do this!

    
## Interpreting results

::: {style="font-size: 0.9em;"}
- Fit the model selected to the validation set and check

    + coefficients - are they similar?
    + significance tests - similar results?
    + MSE - similar prediction error?
    
- If the results are similar between the sets, no big issues with bias from model selection

    + Customary to refit the model to the entire (training + validation) data set
:::

. . .

::: {style="font-size: 0.9em;"}

- If they are quite different, trust the results from the validation set

:::



