---
title: "Inference"
subtitle: "Poisson regression -- Stat 230"
format: 
  revealjs:
    chalkboard: 
      buttons: false
editor: source
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NULL, eval = TRUE, echo = FALSE)

library(ggplot2)
library(ggformula)
library(ggrepel)   # label outliers
library(dplyr)
library(car)
library(gridExtra)
library(broom)
library(Sleuth3)
library(ggthemes)
library(GGally)
library(gt)

puffin <- read.table("http://www.markirwin.net/stat149/Data/puffin.txt", header = TRUE)
colnames(puffin)[1] <- "nests"
puffin <- subset(puffin, nests > 0)

puffin_mod <- glm(nests ~ grass + soil + angle + distance, data = puffin, 
                  family = poisson)
```


## Review: Poisson regression model

- If $Y_i =$ # of events in a fixed interval of time/space, then $Y_i \sim \text{Poisson}(\lambda)$ 


- This leads to the Poisson regression model for counts:

    $$\log(\lambda) = \beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p$$
    or equivalently


## Assumptions

1. **Linearity**: the log of the mean of Y, $\log(\lambda)$, must be a linear function of the predictors

2. **Mean = Variance**: the mean of a Poisson random variable must be equal to its variance

3. **Independence**: the observations must be independent

<!-- ## EDA idea -->

<!-- Plot the log rate vs. predictor variable -->

<!-- Example: Plot the log cancer incidence rate vs. the median age by location -->

<!-- :::: {.columns} -->

<!-- ::: {.column width="60%"} -->
<!-- ```{r} -->
<!-- #| fig-height: 3 -->
<!-- #| fig-width: 5 -->
<!-- #| out-width: 100% -->
<!-- cancer <- cancer |>  -->
<!--   mutate(cancer_incidence = (Cases/Person.years) * 100000) -->
<!-- gf_point(log(cancer_incidence) ~ MedianAge, data = cancer,  -->
<!--          color = ~Location, shape = ~Location) |> -->
<!--   gf_line(linetype = ~Location) + -->
<!--   scale_color_colorblind() + -->
<!--   scale_shape_manual(values = c(1, 16)) + -->
<!--   scale_linetype_manual(values = c(2, 1)) + -->
<!--   labs(x = "Median Age", y = "log cancer incidence rate") -->
<!-- ``` -->
<!-- ::: -->

<!-- ::: {.column width="40%"} -->
<!-- - Is the relationship linear? -->

<!-- - Is the change in the rate with age the same for both groups? -->
<!-- ::: -->
<!-- :::: -->

## Types of inference

1. **Wald based**

    For large enough sample sizes, the sampling distributions for the regression coefficients are approximately normal
    
    $\Longrightarrow$ We can use intro stat-type tests/intervals

. . .

2. **Likelihood based**

    We can use the likelihood function (or deviance) to compare two fitted models
    


## Wald test

:::: {.columns}
::: {.column width="25%"}
**Hypotheses:**
:::

::: {.column width="75%"}
::: {.fragment fragment-index=1}
$H_0: \beta_i = 0 \quad$ vs. $\quad H_a: \beta_i \ne 0$
:::
:::

::::

:::: {.columns}
::: {.column width="25%"}
**Test statistic:**
:::

::: {.column width="75%"}
::: {.fragment fragment-index=2}
$z = \dfrac{\widehat{\beta}_i}{SE(\widehat{\beta}_i)}$
:::
:::

::::

<br>


:::: {.columns}
::: {.column width="25%"}
**Reference distribution:**
:::

::: {.column width="75%"}
::: {.fragment fragment-index=3}
$N(0, 1)$
:::
:::

::::

## Poisson model for puffin nesting data

```{r}
tidy(puffin_mod) |> 
dplyr::mutate(p.value = gtsummary::style_pvalue(p.value)) |>
  gt::gt() |>
  gt::fmt_number(
    columns = c(estimate, std.error, statistic),
    decimals = 3
  ) |>
  gt::tab_options(
    table.width = pct(90),
    table.font.size = px(32),
  ) |>
  gt::cols_width(
    starts_with("term") ~ px(200),
    starts_with("p.") ~ px(200),
    everything() ~ px(200)
  ) |>
  cols_align(
    align = "center",
    columns = p.value
  )
```



## Wald interval


A normal-based confidence interval for ${\beta}_i$

$$\widehat{\beta}_i \pm z^*_{1-\alpha/2} SE(\widehat{\beta}_i)$$

. . .

<br>

But this is on the log scale, so we need to back transform!

$$\left(e^{\widehat{\beta}_i - z^*_{1-\alpha/2} SE(\widehat{\beta}_i)}, \ e^{\widehat{\beta}_i + z^*_{1-\alpha/2} SE(\widehat{\beta}_i)} \right)$$

## Poisson model puffin nesting data {.smaller}

```{r}
tidy(puffin_mod, conf.int = TRUE) |> 
dplyr::mutate(p.value = gtsummary::style_pvalue(p.value)) |>
  gt::gt() |>
  gt::fmt_number(
    columns = c(estimate, std.error, statistic, conf.low, conf.high),
    decimals = 3
  ) |>
  gt::tab_options(
    table.width = pct(98),
    table.font.size = px(28),
  ) |>
  gt::cols_width(
    starts_with("term") ~ px(200),
    starts_with("p.") ~ px(200),
    everything() ~ px(200)
  ) |>
  cols_align(
    align = "center",
    columns = p.value
  )

```

. . .

$−0.027	 \pm 1.96 (0.015)$

. . .

$(e^{-0.0564} \approx `r round(exp(-0.0564), 3)`, e^{-0.0024} \approx `r round(exp(0.0024), 3)`)$

. . .



## Likelihood ratio test {.smaller}

:::: {.columns}
::: {.column width="25%"}
**Full model:** 
:::

::: {.column width="75%"}
$\eta = \beta_0 + \beta_1x_1 + \beta_2x_2 + \cdots + \beta_k x_k + \beta_{k+1} x_{k+1} + \cdots + \beta_p x_p$
:::
::::


:::: {.columns}
::: {.column width="25%"}
**Reduced model: **
:::

::: {.column width="75%"}
$\eta = \beta_0 + \beta_1x_1 + \beta_2x_2 + \cdots + \beta_k x_k$
:::
::::


:::::{.fragment fragment-index=1}

:::: {.columns}
::: {.column width="25%"}
**Hypotheses: **
:::

::: {.column width="75%"}
$H_0: \beta_{k+1} = \cdots = \beta_{p} = 0 \quad \text{vs.} \quad H_a:$ at least one $\beta_j \ne 0$
:::

::::

:::::

:::::{.fragment fragment-index=2}
:::: {.columns}
::: {.column width="25%"}
**Test statistic: **
:::
::::


$$LRT =\text{residual deviance(reduced model)} - \text{residual deviance(full model)}$$
:::::

:::::{.fragment fragment-index=3}

:::: {.columns}
::: {.column width="25%"}
**Reference distribution:** 
:::

::: {.column width="75%"}
$\chi^2$ distribution

d.f. = # $\beta$s in full model $-$ # $\beta$s in reduced model
:::

::::
:::::


## How big is "big enough"?

- Inference requires large enough sample sizes

- This is reasonable if $n$ is "large" 

- OR if $\widehat{\lambda}_i  > 5$
